---
title: "Running a Gillespie simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running a Gillespie simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`origin` grows a cell population in continuous time and replays lineage-recorder
state along the resulting branches. This vignette covers the two entry points
you are most likely to want: simulating a population on its own, and running the
full pipeline that also records barcodes and writes results to disk.

Every code block here has been run as written. The numbers quoted in the text
come from those runs with `seed = 1`; your values will differ if you change the
seed, because a single-founder branching process has very high variance.

```{r library}
library(origin)
```

## Parameters are a plain list

The simulator takes a nested list. In a pipeline that list normally comes from a
JSON file via `jsonlite::fromJSON(path, simplifyVector = FALSE)`, but nothing
requires a file — building it in R is often clearer when you are exploring.

Use `simplifyVector = FALSE` when you do read JSON. The parameter reader expects
scalars to arrive as length-one lists, and the simplifying reader collapses some
of them into vectors that then fail validation.

## Simulating a population

The smallest useful configuration is one cell type and one founder:

```{r population}
params <- list(
  num_init_cells = 1,
  sim_length = list(6),
  random_seed = 1,
  editing_induction = list(timepoint = 0, num_cells = NULL, frac_cells = 1),
  differentiation_induction = list(timepoint = 0, num_cells = NULL, frac_cells = 1),
  cell_type_dict = list(
    founder_cell_type = "progenitor",
    cell_type_params = list(
      progenitor = list(
        cell_cycle_length = 1,
        death_per_cell_cycle_prob = 0.01
      )
    ),
    uninduced_transition_matrix = list(list(1)),
    induced_transition_matrix = list(list(1))
  )
)

population <- simulate_gillespie_population(
  params,
  end_time = 6,
  seed = 1,
  max_cells = 100000L,
  show_progress = FALSE
)
```

`cell_cycle_length` is the *mean* of an exponential division clock, not a fixed
interval, so the division hazard is `1 / cell_cycle_length`.
`death_per_cell_cycle_prob` is a per-cycle probability and is converted
internally to the hazard `-log(1 - p) / cell_cycle_length`. The two are not
interchangeable: anything named `*_hazard` in this package is per unit time,
while anything named `*_prob_per_*` is per cell cycle.

The transition matrices give, for each parent type, the probabilities of each
daughter type. Both daughters are drawn independently, so a division can produce
two of the same fate. With one cell type the matrix is the 1-by-1 `list(list(1))`
above.

The result carries the lineage and a summary of what happened:

```{r inspect}
population$counts
#>          founders         divisions            deaths stochastic_events
#>                 1              1439                16              1455
#>             nodes  surviving_cells
#>              2880             1424

nrow(population$nodes)          #> 2880  every branch segment ever created
nrow(population$terminal_nodes) #> 1424  leaves, including lineages that died
```

`nodes` is the full event-resolved lineage: one row per branch segment, with
`birth_time`, `end_time`, `generation`, `cell_type`, and the flags `is_terminal`,
`alive_at_end` and `died`. Note that `is_terminal` marks a leaf, which includes
cells that died before the horizon. The cells alive at the end are the ones with
`alive_at_end`, and those are usually what you want:

```{r alive}
alive <- population$nodes[population$nodes$alive_at_end %in% c(TRUE, "TRUE"), ]
nrow(alive)  #> 1424
```

## Adding a recorder

`run_gillespie_lineage_pipeline()` grows the population and then replays a
recorder along its branches, writing everything to a directory. Recording needs
more configuration: a barcode sequence to edit, a target layout, and per-state
editing rates.

```{r recorder-params}
editing_state <- list(
  be_mutations_per_target_per_division = 0.02,
  nuc_insertions_per_target_per_division = 0,
  nuc_deletions_per_target_per_division = 0,
  bc_bg_insertion_prob_per_division = 0,
  bc_bg_deletion_prob_per_division = 0,
  bc_substitution_model = "JC",
  bc_sub_model_params = list(0),
  mt_substitution_model = "JC",
  mt_sub_model_params = list(0),
  mt_bg_insertion_prob_per_division = 0,
  mt_bg_deletion_prob_per_division = 0
)

params <- list(
  num_init_cells = 1,
  sim_length = list(6),
  random_seed = 1,
  bc_length = 300,
  max_bc_ints_per_cell = list(2),
  bc_nuc_composition = list(
    frac_a = 0.25, frac_g = 0.25, frac_c = 0.25, frac_t = 0.25
  ),
  be_conversion_pattern = "A --> G",
  be_targets = list(
    num_targets = 8,
    config = "S:10:24",
    edit_rate_class_fractions = list(high = 0.5, medium = 0, low = 0.5),
    editing_window = list(size = 0, decaying = FALSE, close_after_edit = FALSE)
  ),
  editing_induction = list(timepoint = 0, num_cells = NULL, frac_cells = 1),
  differentiation_induction = list(timepoint = 0, num_cells = NULL, frac_cells = 1),
  cell_type_dict = list(
    founder_cell_type = "progenitor",
    cell_type_params = list(
      progenitor = list(
        cell_cycle_length = 1,
        death_per_cell_cycle_prob = 0.01,
        bc_invariant_sites = 0,
        mt_invariant_sites = 0,
        induced_editing_params = editing_state,
        uninduced_editing_params = editing_state
      )
    ),
    uninduced_transition_matrix = list(list(1)),
    induced_transition_matrix = list(list(1))
  )
)
```

Three of these are easy to get wrong, and each is checked:

- `bc_nuc_composition` is required whenever no `barcode_sequence` file is
  supplied. Omitting it, or giving only some of `frac_a`/`frac_g`/`frac_c`/
  `frac_t`, stops the run with a message naming the absent fractions. Fractions
  that cannot form a sequence, such as four zeros, are rejected too.
- `config = "S:10:24"` places targets at position 10 and every 25th base after
  it, so the last of `n` targets sits at `10 + 25 * (n - 1)`. That must fit
  inside `bc_length`. If it does not, the error reports where the last target
  landed, the configured `bc_length`, and a target count that would fit, so you
  do not have to work out which of the four numbers to change.
- `editing_induction` with `num_cells = 1` induces exactly one cell. With a
  single founder that covers the whole population, but with several founders it
  leaves most lineages unedited and every recording downstream is diluted.
  Because nothing fails, this one raises a **warning** rather than an error,
  naming how many founders go uninduced. Use `frac_cells = 1` to induce
  everything. Inducing a subset after time zero is an ordinary design and is not
  warned about.

Building parameters with `modifyList()` is worth avoiding here: it merges nested
lists recursively, so overriding `bc_nuc_composition` with a partial list leaves
the original fractions in place rather than replacing them.

Then run the pipeline:

```{r pipeline}
output_dir <- file.path(tempdir(), "origin_quickstart")
result <- run_gillespie_lineage_pipeline(
  params,
  params_path = NULL,
  output_dir = output_dir,
  overrides = list(progress = FALSE, seed = 1)
)
```

`overrides` takes the same options the `gillespie` block of a JSON parameter file
would carry, so you can set `modalities`, `end_time`, `max_cells` and `seed`
without editing the parameter list. Pass `params_path` when your parameters name
files relative to their own location, such as a `barcode_sequence` or a
prime-editing pegRNA pool; with parameters built in R, `NULL` is correct.

## What it writes

```{r outputs}
list.files(output_dir)
```

The run above produced, among others:

| File | Contents |
| --- | --- |
| `physicell_lineage_full.nwk` | Newick tree over all nodes |
| `physicell_lineage_sampled.nwk` | Newick tree over cells alive at the end (1424 tips) |
| `lineage_nodes.csv.gz` | One row per branch segment, with `generation` and `alive_at_end` |
| `terminal_cells.csv.gz` | The sampled cells |
| `barcode_binary_score_matrix.csv.gz` | Cells by positions, 1 where edited (1424 x 600 here, 3245 edits) |
| `barcode_alleles.csv.gz` | The same shape, carrying allele codes rather than presence |
| `barcode_target_layout.csv.gz` | Which positions are targets, with their edit-rate class |
| `mutation_events.csv.gz` | Every recorded edit, with its time and branch |
| `run_manifest.csv.gz` | The resolved configuration for the run |

The score matrix has one column per (integration, position) named
`int_<i>_pos_<p>`, covering all positions rather than only those observed edited,
so its width is `bc_length` times the number of integrations.

## Choosing this engine

The Gillespie engine draws exact continuous event times, so it has no time step
to tune. The alternative fixed-step path in `legacy/sim5_code.R` discretizes
time with `time_inc`, and its accuracy and cost both depend on that choice:
coarse steps systematically undercount the population, while fine steps cost
proportionally more because every cell is revisited at every step. Prefer this
engine unless you specifically need the legacy path's outputs.

## Where to go next

- `?simulate_gillespie_population` for the full argument list
- `?run_gillespie_lineage_pipeline` for the recording stages and every option
  `overrides` accepts
- `?prepare_physicell_recording_model` for prime editing and PALINCODE, which
  are configured through `prime_editing_backend` and `palincode_adapter` rather
  than `be_targets`
