---
title: "Physical distance versus lineage distance in simulated Visium sections"
author: "remote_mito_clean"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_notebook:
    toc: true
    toc_float: true
    number_sections: true
params:
  input_path: "output/physicell_10000_20260726_171340/visium_spatial"
  max_pairs_per_slice: 50000
  seed: 1
  recent_mrca_hours: 6
---

This parameterized notebook is a reporting stage only: it summarizes tables that
`analyze_physicell_visium.R` has already written and never reruns PhysiCell or
the lineage simulation. `params$input_path` is either one completed
`visium_spatial/` analysis directory or a replicate-batch root (as produced by
`run_physicell_visium_replicates.sh`) under which such directories are found;
each analysis directory must contain `cell_distance_summary.csv`,
`spot_distance_summary.csv` and `slice_summary.csv` (plain or gzipped), and the
pair-level figures additionally use the `cell_pair_lineage_sample.csv[.gz]`
table saved in each `slice_<NNN>/` subdirectory. `params$max_pairs_per_slice`
caps how many saved pairs are loaded per slice for plotting, `params$seed` makes
that subsampling reproducible, and `params$recent_mrca_hours` selects which
`recent_mrca_le_<hours>h_fraction_enrichment` column the enrichment figure
reads. `render_physicell_visium_notebook.R` supplies all four values from its
`--input-path`, `--max-pairs-per-slice`, `--seed` and `--recent-mrca-hours`
options and writes the knitted result to a self-contained HTML file (by default
`<input-path>/physicell_visium_lineage_distance.html`); the notebook itself
produces only figures and tables and writes no files of its own.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 120
)

if(!requireNamespace("ggplot2", quietly = TRUE)){
  stop("The ggplot2 package is required to render this notebook.")
}
library(ggplot2)

theme_set(theme_bw(base_size = 12))
set.seed(as.integer(params$seed))
```

This notebook visualizes the relationship between physical proximity and the
known PhysiCell genealogy after conventional Visium-style sectioning. It can
read either:

1. one completed `visium_spatial/` analysis directory; or
2. a replicate-batch directory created by
   `run_physicell_visium_replicates.sh`.

The dense point plots use at most `r format(params$max_pairs_per_slice, big.mark = ",")`
saved pairs per slice for rendering speed. Binned curves use the complete
`cell_distance_summary.csv.gz` tables produced during analysis. Cell pairs
within one tumor are not independent biological replicates, so uncertainty is
summarized across tumors whenever several replicate runs are supplied.

# Load spatial-lineage output

```{r input-helpers}
#' Resolve a CSV path to whichever of its compressed/uncompressed forms exists
#'
#' Given either `foo.csv` or `foo.csv.gz`, checks both spellings and returns the
#' one present on disk, preferring the gzipped file in both directions.
#'
#' @param path Character scalar path to a CSV, with or without a `.gz`
#'   extension.
#' @return The single existing path as a character scalar. Stops with the list
#'   of candidates that were checked when neither form exists.
resolve_notebook_csv <- function(path){
  candidates <- if(grepl("\\.gz$", path, ignore.case = TRUE)){
    c(path, sub("\\.gz$", "", path, ignore.case = TRUE))
  } else{
    c(paste0(path, ".gz"), path)
  }
  existing <- candidates[file.exists(candidates)]
  if(length(existing) == 0L){
    stop(sprintf("CSV file not found; checked: %s", paste(candidates, collapse = ", ")))
  }
  existing[1]
}

#' Read an analysis CSV that may be stored gzipped
#'
#' Resolves `path` with `resolve_notebook_csv()` and reads the result, opening a
#' text connection first when the resolved file is gzipped. Column names are
#' taken verbatim (`check.names = FALSE`) so the analysis headers survive.
#'
#' @param path Character scalar path to a CSV, with or without a `.gz`
#'   extension.
#' @return A data frame holding the file's rows and unmodified column names.
read_notebook_csv <- function(path){
  input <- resolve_notebook_csv(path)
  if(grepl("\\.gz$", input, ignore.case = TRUE)){
    connection <- gzfile(input, open = "rt")
    on.exit(close(connection), add = TRUE)
    return(utils::read.csv(connection, check.names = FALSE))
  }
  utils::read.csv(input, check.names = FALSE)
}

#' Locate the Visium spatial-lineage analysis directories under an input path
#'
#' Accepts either one completed analysis directory or a replicate-batch root.
#' When `input_path` itself holds `cell_distance_summary.csv[.gz]` it is
#' returned unchanged; otherwise the tree is searched recursively for that
#' summary and only the directories literally named `visium_spatial` are kept,
#' so unrelated per-slice subdirectories are ignored.
#'
#' @param input_path Character scalar directory path; must already exist, since
#'   it is normalized with `mustWork = TRUE`.
#' @return A character vector of normalized analysis-directory paths, sorted
#'   when several replicates are found. Stops when no analysis is found.
discover_visium_analyses <- function(input_path){
  input_path <- normalizePath(input_path, mustWork = TRUE)
  direct_summary <- c(
    file.path(input_path, "cell_distance_summary.csv.gz"),
    file.path(input_path, "cell_distance_summary.csv")
  )
  if(any(file.exists(direct_summary))){
    return(input_path)
  }
  summaries <- list.files(
    input_path,
    pattern = "^cell_distance_summary\\.csv(\\.gz)?$",
    recursive = TRUE,
    full.names = TRUE
  )
  analysis_dirs <- unique(dirname(summaries))
  analysis_dirs <- analysis_dirs[basename(analysis_dirs) == "visium_spatial"]
  if(length(analysis_dirs) == 0L){
    stop(paste(
      "No visium_spatial analysis was found under",
      input_path,
      "Run analyze_physicell_visium.R first."
    ))
  }
  sort(normalizePath(analysis_dirs, mustWork = TRUE))
}

#' Name the tumor replicate that owns an analysis directory
#'
#' @param analysis_dir Character path to a `visium_spatial` analysis directory.
#' @return The basename of that directory's parent, used as the replicate label
#'   in facets, colors, and summary tables.
notebook_replicate_id <- function(analysis_dir){
  basename(dirname(analysis_dir))
}

#' Read a saved cell-pair table, subsampling rows for plotting
#'
#' Rows are drawn without replacement only when the table is larger than
#' `maximum`. The subsample is for rendering speed and affects only the
#' pair-level point/density plots and the descriptive per-section correlations;
#' the binned curves are computed from the complete analysis summaries.
#'
#' @param path Character path to a `cell_pair_lineage_sample.csv[.gz]` table.
#' @param maximum Row cap; tables with at most this many rows are returned
#'   intact.
#' @param seed Integer seed set immediately before sampling, so each slice gets
#'   a reproducible but distinct subsample.
#' @return A data frame with at most `maximum` rows and all original columns.
#' @section Side effects: Calls `set.seed(seed)` when subsampling, which changes
#'   the session's random state.
sample_notebook_pairs <- function(path, maximum, seed){
  values <- read_notebook_csv(path)
  if(nrow(values) > maximum){
    set.seed(seed)
    values <- values[sample.int(nrow(values), maximum), , drop = FALSE]
  }
  values
}
```

```{r load-data}
# Read the cell, spot, and slice summaries plus the per-slice pair samples from
# every discovered analysis directory, tag each row with its replicate/slice,
# bind them into the notebook-wide tables, and print an input overview.
analysis_dirs <- discover_visium_analyses(params$input_path)

pair_tables <- list()
cell_summary_tables <- list()
spot_summary_tables <- list()
slice_summary_tables <- list()
pair_table_index <- 0L

for(analysis_index in seq_along(analysis_dirs)){
  analysis_dir <- analysis_dirs[analysis_index]
  replicate_id <- notebook_replicate_id(analysis_dir)

  cell_summary <- read_notebook_csv(file.path(
    analysis_dir,
    "cell_distance_summary.csv"
  ))
  cell_summary$replicate <- replicate_id
  cell_summary$analysis_dir <- analysis_dir
  cell_summary_tables[[analysis_index]] <- cell_summary

  spot_summary <- read_notebook_csv(file.path(
    analysis_dir,
    "spot_distance_summary.csv"
  ))
  spot_summary$replicate <- replicate_id
  spot_summary$analysis_dir <- analysis_dir
  spot_summary_tables[[analysis_index]] <- spot_summary

  slice_summary <- read_notebook_csv(file.path(
    analysis_dir,
    "slice_summary.csv"
  ))
  slice_summary$replicate <- replicate_id
  slice_summary$analysis_dir <- analysis_dir
  slice_summary_tables[[analysis_index]] <- slice_summary

  pair_paths <- list.files(
    analysis_dir,
    pattern = "^cell_pair_lineage_sample\\.csv(\\.gz)?$",
    recursive = TRUE,
    full.names = TRUE
  )
  for(local_index in seq_along(pair_paths)){
    pair_table_index <- pair_table_index + 1L
    pairs <- sample_notebook_pairs(
      pair_paths[local_index],
      maximum = as.integer(params$max_pairs_per_slice),
      seed = as.integer(params$seed) + 1009L * pair_table_index
    )
    pairs$replicate <- replicate_id
    pairs$slice_id <- basename(dirname(pair_paths[local_index]))
    pairs$section <- paste(replicate_id, pairs$slice_id, sep = "/")
    pair_tables[[pair_table_index]] <- pairs
  }
}

cell_pairs <- if(length(pair_tables) > 0L){
  do.call(rbind, pair_tables)
} else{
  data.frame()
}
cell_distance_summary <- do.call(rbind, cell_summary_tables)
spot_distance_summary <- do.call(rbind, spot_summary_tables)
slice_summary <- do.call(rbind, slice_summary_tables)
rownames(cell_pairs) <- NULL
rownames(cell_distance_summary) <- NULL
rownames(spot_distance_summary) <- NULL
rownames(slice_summary) <- NULL

input_overview <- data.frame(
  analysis_directories = length(analysis_dirs),
  slices = nrow(slice_summary),
  section_cells = sum(slice_summary$section_cells),
  captured_cells = sum(slice_summary$captured_cells),
  occupied_spots = sum(slice_summary$occupied_spots),
  plotted_cell_pairs = nrow(cell_pairs),
  evaluated_cell_pairs = sum(slice_summary$evaluated_cell_pairs),
  stringsAsFactors = FALSE
)
knitr::kable(input_overview, digits = 3)
```

```{r slice-overview}
knitr::kable(
  slice_summary[, c(
    "replicate", "slice_id", "slice_offset_microns", "section_cells",
    "captured_cells", "occupied_spots", "mean_cells_per_occupied_spot",
    "section_capture_fraction", "spearman_spatial_vs_mrca"
  )],
  digits = 3,
  caption = "Sampling and spatial-lineage statistics for each virtual section."
)
```

# Pair-level physical distance versus MRCA age

MRCA age is the elapsed time from terminal sampling back to the most recent
common ancestor. Smaller values indicate closer relatives. The color scale is
logarithmic in the number of plotted cell pairs in each two-dimensional bin.

```{r distance-mrca-density}
if(nrow(cell_pairs) == 0L){
  plot.new()
  title("No captured cell pairs were available")
} else{
  ggplot(
    cell_pairs,
    aes(x = spatial_distance_2d_microns, y = mrca_age_hours)
  ) +
    stat_bin_2d(bins = 55, aes(fill = after_stat(log10(count)))) +
    scale_fill_viridis_c(name = expression(log[10]("plotted pairs"))) +
    labs(
      x = "Projected cell-cell distance (microns)",
      y = "Time back to MRCA (hours)",
      title = "Nearby cells are enriched for recent common ancestry"
    )
}
```

```{r distance-mrca-by-section}
if(nrow(cell_pairs) > 0L){
  ggplot(
    cell_pairs,
    aes(x = spatial_distance_2d_microns, y = mrca_age_hours)
  ) +
    stat_bin_2d(bins = 35, aes(fill = after_stat(log10(count)))) +
    scale_fill_viridis_c(name = expression(log[10]("pairs"))) +
    facet_wrap(~section, scales = "free_x") +
    labs(
      x = "Projected distance (microns)",
      y = "Time back to MRCA (hours)",
      title = "Physical-lineage relationship by tumor and section"
    ) +
    theme(strip.text = element_text(size = 8))
}
```

# Binned lineage correlogram and spatial null

The solid curve is the observed mean MRCA age within each physical-distance
bin. The dashed curve is obtained by permuting lineage tips over fixed spatial
locations. A gap below the null indicates that cells at that spatial scale are
more closely related than expected without spatial lineage structure.

```{r prepare-summary-curves}
# Stack the observed and permutation-null binned MRCA curves into one long
# table keyed by replicate/slice/distance bin, dropping bins whose mean
# distance or mean MRCA age is not finite.
summary_long <- rbind(
  data.frame(
    cell_distance_summary[, c(
      "replicate", "slice_id", "distance_bin",
      "mean_spatial_distance_2d_microns"
    )],
    relationship = "observed",
    mean_mrca_age_hours = cell_distance_summary$mean_mrca_age_hours
  ),
  data.frame(
    cell_distance_summary[, c(
      "replicate", "slice_id", "distance_bin",
      "mean_spatial_distance_2d_microns"
    )],
    relationship = "permuted lineage locations",
    mean_mrca_age_hours = cell_distance_summary$null_mean_mrca_age_hours
  )
)
summary_long$section <- paste(
  summary_long$replicate,
  summary_long$slice_id,
  sep = "/"
)
summary_long <- summary_long[
  is.finite(summary_long$mean_spatial_distance_2d_microns) &
    is.finite(summary_long$mean_mrca_age_hours),
  ,
  drop = FALSE
]
```

```{r observed-null-curves}
ggplot(
  summary_long,
  aes(
    x = mean_spatial_distance_2d_microns,
    y = mean_mrca_age_hours,
    color = relationship,
    linetype = relationship,
    group = interaction(section, relationship)
  )
) +
  geom_line(alpha = 0.45, linewidth = 0.7) +
  geom_point(alpha = 0.55, size = 1.5) +
  scale_color_manual(values = c(
    "observed" = "#2166AC",
    "permuted lineage locations" = "#B2182B"
  )) +
  labs(
    x = "Mean projected distance in bin (microns)",
    y = "Mean time back to MRCA (hours)",
    color = NULL,
    linetype = NULL,
    title = "Observed spatial lineage structure versus permutation null"
  ) +
  theme(legend.position = "bottom")
```

The next curve uses the tumor as the independent unit: sections are averaged
within each tumor first, and the ribbon is plus or minus one standard error
across independently simulated tumors. With one tumor, the mean is shown
without a ribbon.

```{r replicate-level-curve}
# Two-stage aggregation: average the sections within each tumor replicate, then
# average across replicates, so the ribbon is a standard error over
# independently simulated tumors rather than over non-independent cell pairs.
replicate_keys <- unique(summary_long[, c(
  "replicate", "distance_bin", "relationship"
)])
replicate_curve_rows <- vector("list", nrow(replicate_keys))
for(row_index in seq_len(nrow(replicate_keys))){
  selected <-
    summary_long$replicate == replicate_keys$replicate[row_index] &
    summary_long$distance_bin == replicate_keys$distance_bin[row_index] &
    summary_long$relationship == replicate_keys$relationship[row_index]
  replicate_curve_rows[[row_index]] <- data.frame(
    replicate = replicate_keys$replicate[row_index],
    distance_bin = replicate_keys$distance_bin[row_index],
    relationship = replicate_keys$relationship[row_index],
    mean_spatial_distance_2d_microns = mean(
      summary_long$mean_spatial_distance_2d_microns[selected]
    ),
    mean_mrca_age_hours = mean(summary_long$mean_mrca_age_hours[selected]),
    stringsAsFactors = FALSE
  )
}
replicate_curves <- do.call(rbind, replicate_curve_rows)

aggregate_keys <- unique(replicate_curves[, c("distance_bin", "relationship")])
aggregate_curve_rows <- vector("list", nrow(aggregate_keys))
for(row_index in seq_len(nrow(aggregate_keys))){
  selected <-
    replicate_curves$distance_bin == aggregate_keys$distance_bin[row_index] &
    replicate_curves$relationship == aggregate_keys$relationship[row_index]
  lineage_values <- replicate_curves$mean_mrca_age_hours[selected]
  aggregate_curve_rows[[row_index]] <- data.frame(
    distance_bin = aggregate_keys$distance_bin[row_index],
    relationship = aggregate_keys$relationship[row_index],
    mean_spatial_distance_2d_microns = mean(
      replicate_curves$mean_spatial_distance_2d_microns[selected]
    ),
    mean_mrca_age_hours = mean(lineage_values),
    standard_error = if(length(lineage_values) > 1L){
      stats::sd(lineage_values) / sqrt(length(lineage_values))
    } else{
      NA_real_
    },
    tumors = length(lineage_values),
    stringsAsFactors = FALSE
  )
}
aggregate_curves <- do.call(rbind, aggregate_curve_rows)
aggregate_curves <- aggregate_curves[
  order(
    aggregate_curves$relationship,
    aggregate_curves$mean_spatial_distance_2d_microns
  ),
  ,
  drop = FALSE
]

ggplot(
  aggregate_curves,
  aes(
    x = mean_spatial_distance_2d_microns,
    y = mean_mrca_age_hours,
    color = relationship,
    fill = relationship,
    group = relationship
  )
) +
  geom_ribbon(
    aes(
      ymin = mean_mrca_age_hours - standard_error,
      ymax = mean_mrca_age_hours + standard_error
    ),
    alpha = 0.15,
    color = NA,
    na.rm = TRUE
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "observed" = "#2166AC",
    "permuted lineage locations" = "#B2182B"
  )) +
  scale_fill_manual(values = c(
    "observed" = "#2166AC",
    "permuted lineage locations" = "#B2182B"
  )) +
  labs(
    x = "Mean projected distance in bin (microns)",
    y = "Tumor-level mean time back to MRCA (hours)",
    color = NULL,
    fill = NULL,
    title = "Tumor-replicate mean and standard error"
  ) +
  theme(legend.position = "bottom")
```

# Enrichment of recent relatives

```{r recent-relative-enrichment}
# Rebuild the analysis column name for the requested threshold
# (recent_mrca_le_<hours>h_fraction_enrichment, with "." replaced by "_") and
# plot it per section, or report that the analysis saved no such column.
threshold_suffix <- gsub("\\.", "_", format(
  as.numeric(params$recent_mrca_hours),
  trim = TRUE
))
enrichment_column <- paste0(
  "recent_mrca_le_",
  threshold_suffix,
  "h_fraction_enrichment"
)
if(enrichment_column %in% names(cell_distance_summary)){
  enrichment_data <- cell_distance_summary[
    is.finite(cell_distance_summary$mean_spatial_distance_2d_microns) &
      is.finite(cell_distance_summary[[enrichment_column]]),
    ,
    drop = FALSE
  ]
  enrichment_data$enrichment <- enrichment_data[[enrichment_column]]
  enrichment_data$section <- paste(
    enrichment_data$replicate,
    enrichment_data$slice_id,
    sep = "/"
  )
  ggplot(
    enrichment_data,
    aes(
      x = mean_spatial_distance_2d_microns,
      y = enrichment,
      group = section,
      color = replicate
    )
  ) +
    geom_hline(yintercept = 1, color = "grey50", linetype = 2) +
    geom_line(alpha = 0.7) +
    geom_point(size = 1.7) +
    labs(
      x = "Mean projected distance in bin (microns)",
      y = sprintf(
        "Enrichment of MRCA age <= %g hours",
        as.numeric(params$recent_mrca_hours)
      ),
      color = "Tumor replicate",
      title = "Short-range enrichment of recent relatives"
    ) +
    theme(legend.position = "bottom")
} else{
  plot.new()
  title(sprintf(
    "No %g-hour enrichment column was saved; rerun analysis with that threshold",
    as.numeric(params$recent_mrca_hours)
  ))
}
```

# Physical distance versus lineage-edge distance

Division distance is the number of event-tree edges separating two cells via
their common ancestor. This is complementary to elapsed-time branch length
when cell-cycle times vary.

```{r distance-division-density}
division_pairs <- cell_pairs[is.finite(cell_pairs$division_distance), , drop = FALSE]
if(nrow(division_pairs) > 0L){
  ggplot(
    division_pairs,
    aes(x = spatial_distance_2d_microns, y = division_distance)
  ) +
    stat_bin_2d(bins = 55, aes(fill = after_stat(log10(count)))) +
    scale_fill_viridis_c(name = expression(log[10]("plotted pairs"))) +
    labs(
      x = "Projected cell-cell distance (microns)",
      y = "Lineage-edge distance",
      title = "Physical distance versus divisions separating cell pairs"
    )
}
```

# Within-spot and between-spot relationships

Conventional Visium spots contain mixtures of cells. The first plot shows the
cell-pair distribution within one spot versus between distinct spots. The
second gives every sampled spot pair equal weight, rather than allowing spots
with more cells to dominate.

```{r same-spot-pairs}
if(nrow(cell_pairs) > 0L){
  same_spot_data <- transform(
    cell_pairs,
    spot_relationship = ifelse(same_spot, "within the same spot", "different spots")
  )
  ggplot(
    same_spot_data,
    aes(x = spot_relationship, y = mrca_age_hours, fill = spot_relationship)
  ) +
    geom_boxplot(outlier.shape = NA, width = 0.65) +
    coord_cartesian(ylim = stats::quantile(
      same_spot_data$mrca_age_hours,
      c(0, 0.99),
      na.rm = TRUE
    )) +
    scale_fill_manual(values = c(
      "within the same spot" = "#4393C3",
      "different spots" = "#D6604D"
    )) +
    labs(
      x = NULL,
      y = "Time back to MRCA (hours)",
      fill = NULL,
      title = "Lineage relatedness within and between Visium spots"
    ) +
    theme(legend.position = "none")
}
```

```{r spot-distance-curve}
# Keep spot-distance bins that actually sampled spot pairs, rank them by mean
# spot-center distance within each replicate, and plot the equal-weight
# spot-pair mean MRCA age against spot-center distance.
spot_plot_data <- spot_distance_summary[
  spot_distance_summary$spot_pairs > 0L &
    is.finite(spot_distance_summary$mean_mrca_age_hours),
  ,
  drop = FALSE
]
spot_plot_data$spot_distance_order <- ave(
  spot_plot_data$mean_spot_center_distance_microns,
  spot_plot_data$replicate,
  FUN = function(values) rank(values, ties.method = "first")
)
if(nrow(spot_plot_data) > 0L){
  ggplot(
    spot_plot_data,
    aes(
      x = mean_spot_center_distance_microns,
      y = mean_mrca_age_hours,
      group = interaction(replicate, slice_id),
      color = replicate
    )
  ) +
    geom_line(alpha = 0.7) +
    geom_point(aes(size = spot_pairs), alpha = 0.8) +
    scale_size_continuous(name = "Spot pairs", range = c(2, 6)) +
    labs(
      x = "Mean spot-center distance (microns)",
      y = "Equal-spot-pair mean MRCA age (hours)",
      color = "Tumor replicate",
      title = "Visium-resolution spatial lineage relationship"
    ) +
    theme(legend.position = "bottom")
}
```

# Section projection effects

Cells are selected from a finite-thickness three-dimensional slab and then
projected into two dimensions. The following plot reveals when cells that
appear close in the section are farther apart in 3-D.

```{r projection-effects}
if(nrow(cell_pairs) > 0L){
  ggplot(
    cell_pairs,
    aes(
      x = spatial_distance_2d_microns,
      y = spatial_distance_3d_microns - spatial_distance_2d_microns
    )
  ) +
    stat_bin_2d(bins = 50, aes(fill = after_stat(log10(count)))) +
    scale_fill_viridis_c(name = expression(log[10]("plotted pairs"))) +
    labs(
      x = "Projected 2-D distance (microns)",
      y = "3-D minus projected distance (microns)",
      title = "Distance distortion introduced by section projection"
    )
}
```

# Per-section statistics

```{r section-statistics}
# Descriptive per-section statistics over the plotted (possibly subsampled)
# pairs: Spearman correlations of projected distance against MRCA age and
# against lineage-edge distance, plus within-spot pair counts and mean MRCA age.
section_groups <- split(cell_pairs, cell_pairs$section)
section_statistics <- do.call(rbind, lapply(section_groups, function(values){
  data.frame(
    replicate = values$replicate[1],
    slice_id = values$slice_id[1],
    plotted_pairs = nrow(values),
    spearman_distance_vs_mrca = suppressWarnings(stats::cor(
      values$spatial_distance_2d_microns,
      values$mrca_age_hours,
      method = "spearman"
    )),
    spearman_distance_vs_divisions = suppressWarnings(stats::cor(
      values$spatial_distance_2d_microns,
      values$division_distance,
      method = "spearman",
      use = "complete.obs"
    )),
    within_spot_pairs = sum(values$same_spot),
    within_spot_mean_mrca_hours = if(any(values$same_spot)){
      mean(values$mrca_age_hours[values$same_spot])
    } else{
      NA_real_
    },
    stringsAsFactors = FALSE
  )
}))
rownames(section_statistics) <- NULL
knitr::kable(
  section_statistics,
  digits = 3,
  caption = paste(
    "Pair-level correlations are descriptive; tumor replicates, not pairs,",
    "are the independent experimental units."
  )
)
```

# Interpretation

The expected pattern is an increasing mean MRCA age and lineage-edge distance
with increasing physical separation. A strong short-range effect does **not**
imply that every nearby cell pair is closely related: mechanical rearrangement,
growth-boundary interactions, cell loss, and section projection create overlap
between genealogical distances at every spatial scale. The permutation curve
quantifies how much of the observed relationship is lost when lineage labels
are decoupled from cell locations.

For replicated experiments, interpret the between-tumor spread of curves and
statistics. Treating all within-tumor pairs as independent would severely
underestimate uncertainty because each cell participates in many pairs and all
cells share one simulated growth history.
